A Survey of Physical Design Techniques of Information Systems

### 2010-07-26 trunk/cs848-db/Paper.tex

\usepackage{graphics}
\usepackage{color}
\usepackage[pdftex]{hyperref}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{url}
\usepackage{multirow}
\begin{document}

@@

Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database system,
we usually think of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase in
database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
performance. Physical database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed.
Data is stored in the form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and
access the data such that queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing
techniques, partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
techniques \cite{fikelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
where the primary copy of the data resides on disk. This is as opposed to Main Memory Relational Databases (MMDB) where the primary copy of the data resides in
main memory and no disk operations are needed to query the data. Such differences pose new requirements for the physical design of these systems.

@@

Bitmap indices represent another type of index structures that is commonly used and cited in the literature. Using bitmaps as indexes first appeared in the
literature as part of the architecture of the Model 204 commercial database of Computer Corporation of America for the IBM 370 computer \cite{patrick1987model}.
A bitmap index structure is based on a bitmap (i.e. bitset or bit array) data structure where individual bits correspond to values in unique records. Given a
table column $C$ where a record $x$ can have one of six unique values, a traditional bitmap indexe defined for $C$ generates one bitmap vector for each of the
six different values ($B0-B5$). If the value of $C$ for record $x$ is set to 3, the corresponding bit in vector $B3$ is set to 1. Otherwise the bit is set to 0.

Bitmap indices are best used when the set of possible key values in the index is small with a large number of rows, e.g. gender attribute \cite{212001}.
Therefore, bitmap indices outperform B+trees for low cardinality attributes. The reason is that given an attribute with low cardinality, defining a B+tree index
on it means that the tree will have very few leaf nodes, each one storing a very long list of record identifiers (RIDs). On the other hand, defining a
bitmapindex on that same attribute means that such long lists or RIDs will not be stored, rather bit vectors that represent the distinct attribute values will
be stored as previously explained. This leads to a direct benefit of consuming less storage space. Moreover, queries in that case can be answered using bit-wise
logical operatios (e.g. and, or, not) which can improve query performance \cite{253268}. Another advantage of bitmap indices is the ability to perform more
efficient joins by creating bitmap join indices and using those indices to calculate the result of the join \cite{212001}. Although there are many benefits
gained from using bitmap indices, they poorly perform during updates. The reason is that when an updates is to be done to a bitmap index, the whole bit vector
for the index is locked, blocking any access to the index for other operations, then the lock is released when the update is done. That is why bitmap indices
are usually used for read-mostly databases, a famous example of which are data warehouses (Section \ref{SEC-WAREHOUSES}).

Another variation of indexing, a hash table index, often simply known as a hash index is used to organize unordered physical database tables for efficient
access \cite{lightstone2007physical}. This type of index is usually defined for keys that have unique values in the data records. Creating this index requires
passing the attribute (key) value to a hash function that maps it to a starting block address (i.e. bucket address). Afterwards, table insertions will be done
according to the hash function, as well as quering the records later.

@@

Partitioning is an important aspect of physical design as it reduces table scanning time \cite{lightstone2007physical}. There are two main categories of
partitioning: horizontal partitioning and vertical partitioning \cite{agrawal2004integrating}. Horizontal partitioning divides the table into sets of table rows
where each row or record still has all the attributes of the table \cite{ceri1982horizontal, shin1985partitioning}. For example, dividing the data by date where
all data dating less than year 2000 lies in one partition while all data dating more lies in another. On the other hand, vertical partitioning reduces the width
of the dataset by storing some attributes in one table, and some in another. Usually, less frequently queried columns are partitioned together to reduce the
size of the table, especially in data warehouses where not all the columns of a table are analyzed for decision support.

The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance
\cite{lightstone2007physical}. For example, if a table has 1000 records, and a query is only using the first 20 records then having these 20 records in a
partition containing 50 records will save time as opposed to examining all the 1000 records. Horizontal partitioning is more widely used, and is implemented in
all current DBMSes \cite{mysqlpartitioning, sqlsrvpartitioning, sybasepartitioning, postgrepartitioning, oraclepartitioning, db2partitioning} which explains
why there exists many flavors of horizontal paritioning techniques. There are three major types of horizontal partitioning: range partitioning
\cite{dewitt1986gamma}, list partitioning \cite{lightstone2007physical} (which is a variation of range partitioning), and hash partitioning
\cite{lightstone2007physical}. Of course, one can mix and match those partitioning schemes and come up with a composite partitioning that fits a particular
situation \cite{lightstone2007physical}. Table \ref{tab:horpartition} shows different categories of horizontal partitioning, and the situations they are suited
to. Combining partitioning and indexing can often lead to better performance. For example, if List partitioning is used to divide a table by regions and then a
bitmap index is used on that table, finding records in the same region or state can be much faster. Finally, partitioning the data into different physical
devices to achieve parallelism when accessing it can greatly increase the performance of the database.

\begin{table}[!t]
\scalebox{0.7}{
 \begin{tabular}{p{2cm}||p{4cm}|p{4cm}|p{4cm}|p{4cm}}
& \textbf{Range Partitioning} & \textbf{List Partitioning} & \textbf{Hash Partitioning} & \textbf{Composite Partitioning}\\
\hline\hline
\textbf{How?} & If key falls in range & List of keys & Hash function& Hybrid\\
\hline
\textbf{Why?} & Disk limits & Group point values & Range/List are N/A & According to needs\\
\hline
\textbf{Benefits} & \multicolumn{4}{c}{Partition elimination, Improved administration, Fast roll-in/out, and Implicit clustering }\\
\hline
\textbf{Example} & \begin{itemize}
                    \item Zip code 
		    \item Time range
                   \end{itemize} &
		    \begin{itemize}
		    \item Regions
		    \item States
		    \end{itemize} &
		    \begin{itemize}
		    \item EmployeeId
		    \item ProductId
		    \end{itemize} &
		    Hash then List \\
\hline
\textbf{RDBMS} & MySQL, DB2, Oracle, PostgreSQL, SQL Server & MySQL, Oracle, PostgreSQL, SQL Server & MySQL, Oracle & MySQL, Oracle\\
\end{tabular}
}
\caption{Types of horizontal partitioning}
\label{tab:horpartition}
\end{table}

\subsection{Clustering}
Clustering reflects how the records are physically located together on disk. This is very important since sequential I/O access is much cheaper than random
access. Therefore, records that are likely to be queried together should be stored physically together to decrease the number of random I/O accesses.
MOREEEEEEEEEEEEEEEE

@@

A Main Memory Database (MMDB) (sometimes referred to as in-memory database) is one where the primary copy of the data resides in the main memory of the system
rather than on a
disk~\cite{garcia1992main}. It should be noted that this is different from caching. Disk based database systems use the main memory to cache query results.
However, MMDB's primary copy of the data resides in main memory. Accessing data in main memory is faster than disk, and so MMDBs usually have better
performance that disk based systems. However, main memory is more expensive and the size is limited in comparison to disk. However, MMDBs are essential to use
where some of the collected data cannot, or will not, be stored on disk in the first place. For example, phone switch data may be needed for reasoning with
other stored data, but it is captured in real time and will not necessarily be stored on disk. MMDB have many implementation challenges that have been addressed
throughout the research community. These challenges will be discussed throughout this section. %Table~\ref{tab:mmdbsumm} summarizes
the physical design elements of MMDBs.

\begin{table}[t!]
\centering
\scalebox{0.9}{
\begin{tabular}{p{3cm}|p{6cm}|p{6cm}}
\hline
 \textbf{Physical Design Element} & \textbf{Desired Features} & \textbf{Methods/Structures Used}\\
\hline\hline
Index Structures & \begin{itemize}
                    \item No need to store actual values in index \item Larger node size that is aligned with cache line size for cache consciousness
                   \end{itemize} & \begin{itemize}
                                    \item B Tree
\item B+ Tree
\item CSB+ Tree
\item pB+ Tree
\item T Tree
\item CST Tree
\item CSS Tree
                                   \end{itemize}\\
\hline
Materialized Views & \begin{itemize} \item Not highly utilized since processing and memory access is cheap \end{itemize}& N/A\\
\hline
Partitioning & \begin{itemize} \item Only needed in secondary storage to speed up reloading in case of a crash \end{itemize} & \begin{itemize}
                                                                                            \item Horizontal Partitioning
\item Vertical Partitioning
                                                                                           \end{itemize}\\
\hline
Clustering & \begin{itemize} \item Not needed since random or dispersed main memory access is not more expensive than sequential access \end{itemize}
& N/A

\end{tabular}
}
\caption{Summary of physical design of Main Memory Databases (MMDB)}
\label{tab:mmdbsumm}
\end{table}

@@

In this section, we talk about three families of index structures used in MMDB: B Trees, T Trees, and Binary Search Trees. There are more recent index
structures developed for MMDB such as J+ Trees~\cite{luan2009prefetching}, but which have not gained much popularity, and so we do not discuss them in our
survey. %Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB. The T Tree is
the most used index for MMDB (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
indexing. For their purposes, it served as a good index for main memory since it eliminates the need for many key comparisons~\cite{ibmsoliddb}.

 \begin{table}[t!]
\centering
\scalebox{0.6}{
\begin{tabular}{p{2cm}|p{6cm}|p{6cm}}
\hline
 \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
\hline\hline
B Trees &
\begin{itemize}\item Good storage utilization
\item Reasonably quick searching
\item Fast updating
\end{itemize} 
& \begin{itemize}\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
\hline
B+ Trees & 
\begin{itemize}
\item Reasonably quick searching
\item Fast updating
\end{itemize} &
\begin{itemize}
\item Reasonable cache behavior if node fits in cache line\end{itemize}\\
\hline
CSB+ Trees & \begin{itemize}
              \item Same features of B+ Trees
\item Stores child nodes in an array. Stores pointer to first child only and uses index calculation to get the rest
             \end{itemize} &
\begin{itemize}
 \item Good cache performance due to improved locality \& lower tree height (because of larger node size)
\end{itemize}\\
\hline

pB+ Trees & \begin{itemize}
              \item Same features of B+ Trees
             \end{itemize} &
\begin{itemize}
 \item Improves cache miss performance by prefetching more data in each cache miss
\item Node size is also increased leading to the same benefits of CSB+ Trees
\end{itemize}\\
\hline
T Tree &  \begin{itemize}
              \item Contain pointers to data values instead of the values themselves leading to better space utilization
\item Many sorted keys per node
\item Only minimum and maximum keys used for comparisons
\item Contains record pointers in every key
\item Good update \& storage characteristics if Btrees.

             \end{itemize} &
\begin{itemize}
 \item Cache behavior was not considered at time of design
\item Record pointers waste space in cache
\end{itemize}\\
\hline
CST & \begin{itemize}
       \item Binary search tree for maximum value of each node leads to faster search
\item More space efficient by removing pointers \& using array storage. Index calculation used to allocate values
      \end{itemize} &
\begin{itemize}
 \item Node size is aligned with cache line size to avoid misses
\item Higher usage of cached data in binary search tree
\end{itemize}\\
\hline
CSS Trees &  \begin{itemize}
       \item Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node)
\item Mainly suitable for read environments
      \end{itemize} &
\begin{itemize}
 \item Good cache behavior since $m$ is chosen to fit in the cache line size
\end{itemize}\\

@@

Data warehousing has been shown useful in many industries: manufacturing (for order shipment and customer support), retail (for user profiling and inventory
management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management),
telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and healthcare (for outcomes analysis)
\cite{chaudhuri1997overview}.

@@

Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS
databases, and flat files) using Extract, Transfer, and Load (ETL) tools (see Figure \ref{fig:dw}), a data warehouse is usually maintained seperately from the
organization's operational databases \cite{sen2005comparison, chaudhuri1997overview}. This architecture can be justified owing to the fact that operational
databases are finely tuned to satisfy known OLTP requirements and functionalities which are quite different from that of on-line analytical processing (OLAP)
which is supported by data warehousing. Analyzing data for decision support usually requires consolidating data from many heterogeneous sources of varying
quality, or use inconsistent representations, codes and formats. Moreover, understanding trends or making predictions requires historical data, whereas
operational databases store only current data. In OLAP, there's a need to support multidimensional data models and operations which requires special data
organization, access methods, and implementation methods, not generally provided by DBMSs targeted for OLTP \cite{chaudhuri1997overview}. Finally, in data
warehousing, query throughput and response times are more important than transaction throughput which is the major performance metric for operational databases.

@@

Just like all other research areas are moving to the cloud, database systems are moving there too. Accordingly, lots of current research focuses on adapting
storage techniques and index structures to the cloud. For example, Wang et al. propose a new index structure for indexing mult-dimensional data in the
cloud~\cite{wang2010indexing}. Lots of work is currently being done in this area. For example, Wu et al.~\cite{wu2010efficient} propose a modified B+tree index
for datasets stored in each compute node. The exact requirements and solutions for data management in the cloud are still being
researched~\cite{abadi2009data}, and so this is an area that is expected to receive more attention in the future as cloud computing becomes more prevalent.

@@

@@

%\bibliographystyle{IEEEtran}

>>>>

\usepackage{graphics}
\usepackage{color}
\usepackage[pdftex]{hyperref}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{url}
\usepackage{multirow}
\usepackage{array}

\begin{document}

@@

Relational Database Management Systems (RDBMS) have been the most popular type of Information Systems for decades. When we think of designing a database system,
we usually think of which data we need to keep track of and how will we properly organize it into meaningful tables. However, this is not the only phase in
database design. All of the designed tables need to be physically stored on disk. How these tables will be stored and accessed greatly affects the database
performance. Physical database design is concerned with the actual storage of data on disk. This means how the files are organized, and how they are accessed.
Data is stored in the form of records in files which reside on a physical disk. In this sense, we try to find the most efficient way possible to store and
access the data such that queries are answered and executed efficiently. In the traditional relational database system, this involved selecting proper indexing
techniques, partitioning the data for more efficient access, materializing the results of some queries, clustering and compressing the data, and many other
techniques \cite{finkelstein1988physical,lightstone2007physical}. The physical design aspects of relational databases have been studied in much detail. However,
over time, different types of Information Systems have emerged that have different characteristics. For example, most relational databases have been disk-based
where the primary copy of the data resides on disk. This is as opposed to Main Memory Relational Databases (MMDB) where the primary copy of the data resides in
main memory and no disk operations are needed to query the data. Such differences pose new requirements for the physical design of these systems.

@@

Bitmap indices represent another type of index structures that is commonly used and cited in the literature. Using bitmaps as indexes first appeared in the
literature as part of the architecture of the Model 204 commercial database of Computer Corporation of America for the IBM 370 computer \cite{patrick1987model}.
A bitmap index structure is based on a bitmap (i.e. bitset or bit array) data structure where individual bits correspond to values in unique records. Given a
table column $C$ where a record $x$ can have one of six unique values, a traditional bitmap index defined for $C$ generates one bitmap vector for each of the
six different values ($B0-B5$). If the value of $C$ for record $x$ is set to 3, the corresponding bit in vector $B3$ is set to 1. Otherwise the bit is set to 0.

Bitmap indices are best used when the set of possible key values in the index is small with a large number of rows, e.g. gender attribute \cite{212001}.
Therefore, bitmap indices outperform B+trees for low cardinality attributes. The reason is that given an attribute with low cardinality, defining a B+tree index
on it means that the tree will have very few leaf nodes, each one storing a very long list of record identifiers (RIDs). On the other hand, defining a
bitmap index on that same attribute means that such long lists or RIDs will not be stored, rather bit vectors that represent the distinct attribute values will
be stored as previously explained. This leads to a direct benefit of consuming less storage space. Moreover, queries in that case can be answered using bit-wise
logical operations (e.g. and, or, not) which can improve query performance \cite{253268}. Another advantage of bitmap indices is the ability to perform more
efficient joins by creating bitmap join indices and using those indices to calculate the result of the join \cite{212001}. Although there are many benefits
gained from using bitmap indices, they poorly perform during updates. The reason is that when an updates is to be done to a bitmap index, the whole bit vector
for the index is locked, blocking any access to the index for other operations, then the lock is released when the update is done. That is why bitmap indices
are usually used for read-mostly databases, a famous example of which are data warehouses (Section \ref{SEC-WAREHOUSES}).

Another variation of indexing, a hash table index, often simply known as a hash index is used to organize unordered physical database tables for efficient
access \cite{lightstone2007physical}. This type of index is usually defined for keys that have unique values in the data records. Creating this index requires
passing the attribute (key) value to a hash function that maps it to a starting block address (i.e. bucket address). Afterwards, table insertions will be done
according to the hash function, as well as querying the records later.

@@

Partitioning in physical database design is a method for reducing the workload on any one hardware component, like an individual disk, by partitioning
(dividing) the data over several disks \cite{lightstone2007physical}. There are two main categories of partitioning: horizontal partitioning and vertical
partitioning \cite{agrawal2004integrating}. Horizontal partitioning divides the table into sets of table rows where each row or record still has all the
attributes of the table \cite{ceri1982horizontal, shin1985partitioning}. For example, dividing the data by date where all data dating less than year 2000 lies
in one partition while all data dating more lies in another. On the other hand, vertical partitioning reduces the width of the dataset by storing some
attributes in one table, and some in another. Usually, less frequently queried columns are partitioned together to reduce the size of the table, especially in
data warehouses where not all the columns of a table are analyzed for decision support.

The main advantage of any type of partitioning is to reduce the amount of time it takes to scan a table which in turn improves performance
\cite{lightstone2007physical}. For example, if a table has 1000 records, and a query is only using the first 20 records then having these 20 records in a
partition containing 50 records will save time as opposed to examining all the 1000 records. Horizontal partitioning is more widely used, and is implemented in
all current DBMSes \cite{mysqlpartitioning, sqlsrvpartitioning, sybasepartitioning, postgrepartitioning, oraclepartitioning, db2partitioning} which explains
why there exists many flavors of horizontal partitioning techniques. There are three major types of horizontal partitioning: range partitioning
\cite{dewitt1986gamma}, list partitioning \cite{lightstone2007physical} (which is a variation of range partitioning), and hash partitioning
\cite{lightstone2007physical}. Of course, one can mix and match those partitioning schemes and come up with a composite partitioning that fits a particular
situation \cite{lightstone2007physical}. Table \ref{tab:horpartition} shows different categories of horizontal partitioning, and the situations they are suited
to. Combining partitioning and indexing can often lead to better performance. For example, if List partitioning is used to divide a table by regions and then a
bitmap index is used on that table, finding records in the same region or state can be much faster. Finally, partitioning the data into different physical
devices to achieve parallelism when accessing it can greatly increase the performance of the database.

\begin{table}[!t]
\centering
\scalebox{0.7}{
 \begin{tabular}{p{1.8cm}||p{4cm}|p{3.6cm}|p{4cm}|p{5cm}}
& \textbf{Range Partitioning} & \textbf{List Partitioning} & \textbf{Hash Partitioning} & \textbf{Composite Partitioning}\\
\hline\hline
\textbf{How?} & If key falls in range & List of keys & Hash function& Hybrid\\
\hline
\textbf{Why?} & Disk limits & Group point values & Range/List are N/A & According to needs\\
\hline
\textbf{Benefits} & \multicolumn{4}{p{16cm}}{\centering Partition elimination \linebreak Improved administration \linebreak Fast roll-in/out \linebreak Implicit
clustering }\\
\hline
\textbf{Example} & %\begin{itemize} 
                    \raggedright - Zip code \linebreak
		    - Time range &
                  \raggedright
		    - Regions \linebreak
		    - States
		     & \raggedright
		    - EmployeeId \linebreak
		    - ProductId &
		    Hash then List \\
\hline
\textbf{RDBMS} & MySQL, DB2, Oracle, PostgreSQL, SQL Server & MySQL, Oracle, PostgreSQL, SQL Server & MySQL, Oracle & MySQL, Oracle\\
\end{tabular}
}
\caption{Types of horizontal partitioning}
\label{tab:horpartition}
\end{table}

\subsection{Clustering}
Clustering is the process of grouping related data together on disk for efficiency of access and maximizing resource utilization. Data clustering benefits from
the fact that sequential I/O access in physical disks is much cheaper than random access. Therefore, records that are likely to be queried together are
clustered (stored physically together) to decrease the number of random I/O accesses. Whenever the number of I/O random accesses is decreased, CPU blocks are
also reduced which result in an overall reduction of CPU costs improving the query performance, especially for multidimensional queries (e.g. group by)
\cite{lightstone2007physical}. The same concept of data clustering can similarly be applied to multiple dimensions (columns or column attributes) of a table.
Kennedy et al. \cite{kennedy2005} summarizes the benefits of multidimensional clustering as follows:

\begin{itemize}
 \item Dramatic reduction in I/O requirements due to clustering.
 \item Scans on any dimension index provide clustered data access.
 \item Dimensions/slices can be accessed independently using their block indexes.
 \item Block index scans can be combined (ANDed, ORed).
 \item Access to the clustered data is much faster.
\end{itemize}

On the other hand, defining clustering keys, clustering scheme, and the granularity of clustering which makes the problem of designing an optimal clustering in
this context has combinatorial complexity, having an extremely large search space \cite{jagadish1999snakes}.

@@

A Main Memory Database (MMDB) (sometimes referred to as in-memory database) is one where the primary copy of the data resides in the main memory of the system
rather than on a
disk~\cite{garcia1992main}. It should be noted that this is different from caching. Disk based database systems use the main memory to cache query results.
However, MMDB's primary copy of the data resides in main memory. Accessing data in main memory is faster than disk, and so MMDBs usually have better
performance that disk based systems. However, main memory is more expensive and the size is limited in comparison to disk. However, MMDBs are essential to use
where some of the collected data cannot, or will not, be stored on disk in the first place. For example, phone switch data may be needed for reasoning with
other stored data, but it is captured in real time and will not necessarily be stored on disk. MMDB have many implementation challenges that have been addressed
throughout the research community. These challenges will be discussed throughout this section.

@@

In this section, we talk about three families of index structures used in MMDB: B Trees, T Trees, and Binary Search Trees. There are more recent index
structures developed for MMDB such as J+ Trees~\cite{luan2009prefetching}, but which have not gained much popularity, and so we do not discuss them in our
survey. Table~\ref{tab:mmdbindexsumm} summarizes the features of the different index structures discussed in this section with respect to MMDB. The T Tree is
the most used index for MMDB (E.g. MySQL CLuster). However, there are other indexes used as well. For example, IBM's solidDB uses Trie (or prefix tree) for its
indexing. For their purposes, it served as a good index for main memory since it eliminates the need for many key comparisons~\cite{ibmsoliddb}.

 \begin{table}[t!]
\centering
\scalebox{0.8}{
\begin{tabular}{p{2.2cm}|p{6cm}|p{6cm}}
\hline
 \centering \textbf{Index Structure} & \textbf{Features} & \textbf{Cache Consciousness}\\
\hline\hline
B Trees & 
\raggedright -Good storage utilization \linebreak  \linebreak
\raggedright -Reasonably quick searching \linebreak \linebreak
-Fast updating 
& 
-Reasonable cache behavior if node fits in cache line\\
\hline
B+ Trees &
\raggedright -Reasonably quick searching \linebreak \linebreak
\raggedright -Fast updating
&
-Reasonable cache behavior if node fits in cache line\\
\hline
CSB+ Trees & 
-Same features of B+ Trees \linebreak \linebreak
-Stores child nodes in an array. Stores pointer to first child only and uses index calculation to get the rest
 &
-Good cache performance due to improved locality \& lower tree height (because of larger node size)\\
\hline

pB+ Trees & 
-Same features of B+ Trees
             &
-Improves cache miss performance by prefetching more data in each cache miss \linebreak \linebreak
-Node size is also increased leading to the same benefits of CSB+ Trees\\
\hline
T Tree &  
-Contain pointers to data values instead of the values themselves leading to better space utilization \linebreak \linebreak
-Many sorted keys per node \linebreak \linebreak
-Only minimum and maximum keys used for comparisons \linebreak \linebreak
\raggedright -Contains record pointers in every key \linebreak \linebreak
-Good update \& storage characteristics if Btrees.
             &
-Cache behavior was not considered at time of design \linebreak \linebreak
-Record pointers waste space in cache\\
\hline
CST & 
-Binary search tree for maximum value of each node leads to faster search \linebreak \linebreak
-More space efficient by removing pointers \& using array storage. Index calculation used to allocate values &
-Node size is aligned with cache line size to avoid misses \linebreak \linebreak
-Higher usage of cached data in binary search tree\\
\hline
CSS Trees & 
-Fast traversal in $log_{m+1}n$ ($m$ is the number of keys per node) \linebreak \linebreak
-Mainly suitable for read environments &
-Good cache behavior since $m$ is chosen to fit in the cache line size\\

@@

Data warehousing has been shown useful in many industries: manufacturing (for order shipment and customer support), retail (for user profiling and inventory
management), financial services (for claims analysis, risk analysis, credit card analysis, and fraud detection), transportation (for fleet management),
telecommunications (for call analysis and fraud detection), utilities (for power usage analysis), and health care (for outcomes analysis)
\cite{chaudhuri1997overview}.

@@

Although data in a data warehouse is extracted and loaded from multiple on-line transaction processing (OLTP) data sources (including DB2, Oracle, IMS
databases, and flat files) using Extract, Transfer, and Load (ETL) tools (see Figure \ref{fig:dw}), a data warehouse is usually maintained separately from the
organization's operational databases \cite{sen2005comparison, chaudhuri1997overview}. This architecture can be justified owing to the fact that operational
databases are finely tuned to satisfy known OLTP requirements and functionalities which are quite different from that of on-line analytical processing (OLAP)
which is supported by data warehousing. Analyzing data for decision support usually requires consolidating data from many heterogeneous sources of varying
quality, or use inconsistent representations, codes and formats. Moreover, understanding trends or making predictions requires historical data, whereas
operational databases store only current data. In OLAP, there's a need to support multidimensional data models and operations which requires special data
organization, access methods, and implementation methods, not generally provided by DBMSs targeted for OLTP \cite{chaudhuri1997overview}. Finally, in data
warehousing, query throughput and response times are more important than transaction throughput which is the major performance metric for operational databases.

@@

Just like all other research areas are moving to the cloud, database systems are moving there too. Accordingly, lots of current research focuses on adapting
storage techniques and index structures to the cloud. For example, Wang et al. propose a new index structure for indexing multi-dimensional data in the
cloud~\cite{wang2010indexing}. Lots of work is currently being done in this area. For example, Wu et al.~\cite{wu2010efficient} propose a modified B+tree index
for datasets stored in each compute node. The exact requirements and solutions for data management in the cloud are still being
researched~\cite{abadi2009data}, and so this is an area that is expected to receive more attention in the future as cloud computing becomes more prevalent.

@@

\begin{table}[!t]
\centering
\scalebox{0.7}{
\begin{tabular}{p{2cm}||p{5cm}|p{4cm}|p{4cm}|p{4cm}}
 & \textbf{Index Structures} & \textbf{Materialized Views} & \textbf{Partitioning} & \textbf{Clustering}\\
\hline\hline
\textbf{RDBMS}& E.g. B+trees, Bitmap Index, Hash Tree & Store results of common complicated queries on disk & Horizontal Partitioning (range, list, hash,
composite) Vertical Partitioning& Group related items together on disk\\
\hline
\textbf{MMDB} & 
-No need to store actual data values in index \linebreak \linebreak 
-Larger node size that is aligned with cache line size \linebreak \linebreak
-E.g. B+tree, CSB+tree, T tree, CST tree & Not highly utilized since processing and main memory access is cheap & Only needed in secondary storage to speed up
reloading in case of a crash\linebreak \linebreak 
-Horizontal \& Vertical partitioning& 
Not needed since random access in main memory costs the same as sequential access\\
\hline
\textbf{XML DB} & Need a number scheme. E.g. B+tree and variations used. & XQuery and XPath results are materialized & Horizontal partitioning based on inlining
or node
type & Based on document order or structural similarity\\
\hline
\textbf{Data Warehouses} & E.g. Universal B+tree, Bitmap index, Projection Index & 
-Essential to improve performance \linebreak \linebreak 
-Not applicable to adhoc queries & 
-Horizontal partitioning based on time dimension \linebreak \linebreak 
-Vertical partitioning of seldom used columns. & Data is
clustered by
nature.\\
\end{tabular}
}
\caption{Summary of the different elements of physical design in the studied systems}
\label{tab:summary}
\end{table}

@@

In this survey, we present four elements of physical design in database: index structures, materialized views, partitioning, and clustering. We started out
by explaining how these design factors are implemented in traditional Relational Database Management Systems (RDBMS), and then presented how they are used in
other information systems. We focused on three types of information systems: Main Memory Database Systems (MMDB), XML Databases, and Data Warehouses. We
presented the challenges involved with each type of system, and how each of the four physical design elements were adapted or utilized to overcome these
challenges. Table~\ref{tab:summary} summarizes the results of our survey. From our literature review, we found that the B+tree is the most utilized index
structure, and has been adapted and customized to fit the different needs of other systems. For example, it was made more cache conscious to fit MMDB needs,
and was changed to apply to multidimensions to fit the Data Warehouses needs. Materialized views were found to be heavily utlitized in all the systems we
presented, except in MMDBs. Partitioning was also used in all systems, but was only used for data reloading in MMDBs. Horizontal partitioning is much more
common that vertical partitioning. Clustering is very important in all systems, except for MMDBs. Clustering is especially very important for Data Warehouses
where huge amounts of data are stored.

It was interesting to see how the different nature of these systems affects the physical design. Throughout the survey, we have showed some of the
commonalities between the different systems. Moreover, understanding how to meet these needs would help us design mash-ups of these different systems. For
example, just like relational database systems shifted into main memory, we might have an XML main memory database system. In that case, the index structures
fitted for XML data indexing will have to be adapted for main memory by making sure they are aligned with the cache line size, for example. Similarly, if XML
data warehouses are developed, more elaborate path indexing techniques and more clustering would be required.
